\documentclass{easychair}

\usepackage{booktabs}
\usepackage{amsmath, amssymb}
\usepackage{hyperref}
\usepackage{listings}

\title{Bidirectionality vs.\ the Corrig\'e: \\
A Per-PASS, Per-Attempt, and Myers-Transcription Audit of Two SOTA Lean Provers}

\author{David Appadourai}
\institute{Universit\`a di Roma Tor Vergata, Italy \and IHPST, France}
\authorrunning{Appadourai}
\titlerunning{Bidirectionality vs.\ the Corrig\'e}

\begin{document}

\maketitle

\section{The metric}

Pass rate measures one thing: whether the kernel accepts the script. Bidirectionality measures another: whether a Lean-literate mathematician can reconstruct the proof's case structure, named lemmas, and strategy from the script alone. The companion paper~\cite{appadourai-wdwft} names this property and argues that the bipartite ontology cannot articulate it. With a paired corrig\'e, the criterion becomes per-PASS computable. Each named corrig\'e move appears in the script as a named operation (\texttt{have h\,:\,T := proof}, \texttt{apply <named\_lemma>}) or is absorbed into a closing automation chain (\texttt{nlinarith [shotgun]}, \texttt{<;> field\_simp <;> ring\_nf}). Three discrimination levels emerge: \emph{(i)} move-level inscription (each corrig\'e move appears as \texttt{have}); \emph{(ii)} tactic-level inscription (closure by named lemma vs.\ \texttt{fun\_prop} / \texttt{<;>} cascade); \emph{(iii)} inscription-uniqueness (one \texttt{have} per move). Score per PASS: (a) route-preserving, (b) route-equivalent at lower abstraction, (c) opaque automation.

\section{Setup}

Corpus: agr\'egation interne 2005--2013~\cite{mercier-rombaldi}, a French print-only exam absent from Numina, Lean-Workbook, miniF2F, PutnamBench, ProofNet, AOPS --- verified by lemma-signature grep on Lean-Workbook (140K rows, 0 hits), full-text and substring matches against the released portions of the others (0 hits). From 412 audited sub-questions, 20 focal theorems autoformalised in Lean 4 (Mathlib v4.30.0-rc2): 7 algebra, 13 analysis. Selection: pre-experiment 4-band stratification at $K{=}8$ pilot crossed with algebra/analysis balance, model outputs not inspected pre-selection. Provers: Goedel-Prover-V2-32B (RL closure-tuning,~\cite{goedelv2}) and Kimina-Prover-72B (SFT informal-formal alignment via \texttt{<think>} blocks,~\cite{kimina}). Both arms ran at $K = 64$, max\_tokens $= 32{,}000$ matched to each model's published protocol. Total: 2,560 attempts, 805 PASSes (520 Goedel + 285 Kimina). Per-PASS classification was hand-mediated through Sonnet sub-agents calibrated on representative samples.

\section{Findings}

The bidirectional gap is signed, quantitative, and concentrated. \emph{First,} per-PASS classification of all 805 verified scripts yields Kimina 147 (a) / 137 (b) / 1 (c) and Goedel 102 (a) / 418 (b) / 0 (c). Kimina preserves the corrig\'e route at 51.6\% per PASS, Goedel at 19.6\%, despite Goedel's 1.83$\times$ higher PASS rate (520 vs.\ 285). The gap concentrates in seven theorems where Kimina's outputs name the corrig\'e move explicitly and Goedel's outputs absorb it into a closing automation chain.

\emph{Second,} per-attempt aggregates expose a paradigm signature invisible at the PASS layer. Goedel-RL always commits to a Lean script (1280 attempts, 1280 with a \texttt{lean4} block, 0 truncations to no-block). Kimina-SFT does not: 159 of 1280 attempts (12.4\%) emit no usable script --- 50 are 0-character outputs, 110 have unclosed \texttt{<think>} tags overrunning the 32K budget. Goedel-V2-32B commits to a script in 1280/1280 attempts. Kimina-72B commits in 1121/1280. Whether the locus is training paradigm, scale (32B vs.\ 72B), or \texttt{<think>}-decoder is left for ablation.

\emph{Third,} the showcase theorem P12 ($x^2+xy+y^2 \geq (3/4)x^2$ via $(x/2+y)^2 \geq 0$) inverts the arm-quality ranking. Per-PASS hand-classification: Goedel 55/56 (a), Kimina 9/63 (a). Earlier regex-based estimates counted \texttt{(x/2+y)} occurrences inside \texttt{nlinarith} hint-lists as route-preserving; the per-PASS audit shows these are the (b) shotgun-closure pattern. The arm-quality ranking is per-theorem, not constant.

\section{The two ceilings}

Two theorems produce 0 PASSes in 128 attempts each: P11.subq\_III\_1 ($g$ differentiable on $]p,q[$ with $|g'| \leq M \Rightarrow g$ has a finite limit at $q^-$; two-line corrig\'e via MVT + Cauchy criterion) and P9.subq\_22 (winding number invariance under reparameterisation). The single Kimina PASS originally listed for P9.subq\_22 is a Lake false-positive: the script body terminates at \texttt{:= by} with no closing tactic. \texttt{lake env lean} in batch mode accepts the malformed proof, while an interactive lean-lsp re-check on the same Mathlib v4.30.0-rc2 environment rejects it. The two false-positives reported below are pipeline-level artifacts of this kind, not Lean kernel false-positives.

To localise the failures, a Claude Opus 4.7 + lean-lsp-mcp transcription was produced in the Myers role (faithful transcription of corrig\'e moves into named operations) for all 20 theorems, with proof-state inspection, multi-tactic probing, and Mathlib API search through interactive lean-lsp. Of 20 baseline scripts, 19 reach (a) on first or second iteration. P11.subq\_III\_1 closes in one iteration via \texttt{Convex.lipschitzOnWith\_of\_nnnorm\_hasDerivWithin\_le} (MVT) and \texttt{cauchy\_map\_iff\_exists\_tendsto} (Cauchy criterion). The Mathlib API is complete. The failure is premise-selection. P9.subq\_22 also bounces the baseline, but with a structured diagnosis: the Lean statement does not fix $\varphi(0) = 0$ (the corrig\'e silently does), and the period-shift invariance of the rotation index (Q21 of the original problem) is absent from Mathlib v4.30.0-rc2. \emph{In short,} the first ceiling is provers' weakness, the second a content gap. \emph{In effect,} the Claude+lean-lsp-mcp Myers transcription functions as a diagnostic instrument that ML timeout cannot.

\section{Limitations}

The score is corpus-relative. Whether the pattern survives in other traditions is open. The $K{=}64$ main run is single-seed for the per-PASS classification. On a Goedel 3-theorem stratified subset (concrete, standard, abstract), two secondary vLLM seeds (12345, 67890) reproduce per-PASS route rates within $\leq 3.2$\,pp (P12: 75/76/73\%, P11.I.3.b: 0/0/0\%, P7: 0/0/0\%). Kernel-PASS rates vary by 9--14\,pp. The bidirectional ratio is seed-stable on this subset. Full multi-seed coverage and Kimina re-seeding deferred. Per-PASS classification is sub-agent-mediated, in the same sense that referee judgment of readability is. A 3-rater protocol (Sonnet original, Gemini 3.1 Pro, GPT-5.5 --- three independent model families) achieves Fleiss' $\kappa = 0.87$ across 52 valid triples. Pairwise: Sonnet--Gemini $\kappa = 0.81$, Sonnet--GPT-5.5 $\kappa = 0.81$, Gemini--GPT-5.5 $\kappa = 1.00$. Excluding P16.subq\_IV\_3 (slope/multiplicative bound), all three raters agree on every PASS (n=37). The (a)/(b) boundary is genuinely fuzzy on that one criterion. The framework is unanimous across model lineages elsewhere. All 805 PASSes were verified via \texttt{lake env lean} on Mathlib v4.30.0-rc2. Two Lake false-positives surfaced under interactive lean-lsp re-verification (the truncated P9.subq\_22 PASS, and a P15.subq\_I\_2 Kimina PASS invoking a lemma name absent from current Mathlib). Both are Kimina, both remain in the 805 denominator, and their exclusion shifts Kimina's per-PASS route rate by less than 0.5\,pp. The main-run false-positive rate is bounded below by 2/805. The Myers transcription is Claude-mediated. Comparison to human-written Mathlib tactic scripts remains future work.

\smallskip
Code, data, verdicts, per-(theorem, prover) classifications, and Claude-mediated Myers transcriptions: \url{https://github.com/Pythiagora/lean-prover-bidirectionality} (MIT; corrig\'es not redistributed).

\begin{thebibliography}{9}
\bibitem{appadourai-wdwft} Appadourai, D. (in submission). \emph{Why Do We Formalize Theorems? Informal Proof, Tactic Script, Proof Term: A Three-Level Analysis of Mathematical Proof.}
\bibitem{dsp} Jiang, A.\ Q.\ et al. (2022). \emph{Draft, Sketch, and Prove.} arXiv:2210.12283.
\bibitem{leanstar} Lin, H.\ et al. (2024). \emph{Lean-STaR.} arXiv:2407.10040.
\bibitem{goedelv2} Lin, Y.\ et al. (2025). \emph{Goedel-Prover-V2.} arXiv:2508.03613.
\bibitem{mercier-rombaldi} Mercier, D.-J., Rombaldi, J.-\'E. (2013). \emph{Annales de l'agr\'egation interne 2005--2013.} Publibook, ISBN 978-2-342-00940-8.
\bibitem{kimina} Moonshot AI (2025). \emph{Kimina-Prover Preview.} arXiv:2504.11354.
\bibitem{leanworkbook} Ying, R.\ et al. (2024). \emph{Lean Workbook.} arXiv:2406.03847.
\end{thebibliography}

\end{document}
